A boundary integral equation approach to computing eigenvalues of the Stokes operator

### 2019-04-10 doc/draft-01-stokes/04numerical.tex
update manuscript

@@

@@

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_barbell_001_sings_plot}
    \caption{Smallest singular value corresponding to the
    first 119 computed eigenfunctions of the barbell-shaped
    domain.}
    \label{subfig:barbell_sings}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_barbell_001_coeffs_plot}
    \caption{Normalized Chebyshev coefficients of $f^N(k)$ on
      3 different intervals in $k$.}
    \label{subfig:barbell_coeffs}
  \end{subfigure}
  \caption{Diagnostics for the barbell eigenvalues.}
  \label{fig:barbell_diagnostics}
\end{figure}

We consider the barbell-shaped domain in \cref{fig:barbell}.
This domain is the union of a square of side-length 6,
a square of side-length 3, and a ``bridge'' connecting
them of height 1 and width 5/2.
%
For simplicity, we round the corners of the domain
to obtain a smooth object.
%
Applying the approach described in~\cite{epstein2016smoothed},
the corners of the domain are rounded using the Gaussian kernel
\begin{equation}
  \nonumber
\phi(x) = \frac{1}{\sqrt{2\pi h}} e^{-x^2/(2 h^2)} \, ,
\end{equation}
with $h\approx 0.06$. This leaves the domain unperturbed
to high precision outside of a radius of $0.1$ around
each corner.
%
The eigenfunctions of such a domain display the well-known
localization property~\cite{trefethen2006computed}:
many of the eigenfunctions are approximately supported
within one of the squares.
%
We compute these eigenfunctions corresponding to
eigenvalues $k^2$ with $k$ in the range
$0.5 \leq k \leq 6.5$.

%
The panels are divided adaptively so that the smallest
panels in the rounded corner are smaller than $10^{-2}$.
%
This results in $N_p = 412$ after enforcing the
level-restriction property described in
\cref{rmk:levelrestrict}
and enforcing that no panel is larger than
one wavelength for the largest $k$
(here $\lambda=2\pi/6.5$).

As this is a simply-connected domain,
the eigenvalues are estimated by finding the values
$k$ for which $\cI-2\cDk-2\cW$ is non-invertible.
%
Let $f^N(k) = \det (\cI^N-2\cDk^N-2\cW^N)$.
To find the roots of $f^N(k)$, we fit a \texttt{chebfun}
representation of $f^N(k)$ on each of the intervals
$[j/2,(j+1)/2]$ for $j = 1,\ldots,12$.
%
We plot the absolute value of the Chebyshev coefficients
of $f^N(k)$ on the intervals $[0.5,1.0]$, $[3.0,3.5]$,
and $[6.0,6.5]$ in \cref{subfig:barbell_coeffs}.
%
As expected, the coefficients decay exponentially
to zero, with more terms required at higher
frequencies.
%

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig/many_inclusions_gallery}
  \caption{Vorticity plots of...}
  \label{fig:many_inclusions_gallery}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_sings_plot}
    \caption{Smallest singular value corresponding to the
    first.}
    \label{subfig:many_inclusions_sings}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_coeffs_plot}
    \caption{Normalized Chebyshev coefficients of $f^N(k)$ on
      3 different intervals in $k$.}
    \label{subfig:many_inclusions_coeffs}
  \end{subfigure}
  \caption{Diagnostics for the ... eigenvalues.}
  \label{fig:many_inclusions_diagnostics}
\end{figure}

@@

%
For this smooth shape, the panels are chosen
adaptively so that the coordinates, the first
and second derivatives of the coordinates, and
the arc-length function are resolved to a precision
of $10^{-3}$ on each panel.
%
This results in $N_p = 224$ after enforcing the
level-restriction property described in
\cref{rmk:levelrestrict}
and enforcing that no panel is larger than
one wavelength for the largest $k$
(here $\lambda=2\pi/9$).

@@

We compute the roots of these Chebyshev interpolants
and apply the post-processing described above.
%
There were 103 total roots: 21 were removed because
the imaginary part was too large and 6 pairs were found
with values within $\sqrt{\epscheb}$ of each other.
%
For these 6 pairs, none represented two distinct
eigenvalues or a double root.
%
This leaves 76 roots in the range $0\leq k \leq 9$.
%
We plot the smallest singular value of
$\cI^N-\cDk^N-2i\cSk^N-2\cW^N$ for each of these roots in
\cref{subfig:many_inclusions_sings}
and plot the vorticity of the eigenfunctions
in \cref{fig:many_inclusions_gallery}.
%
The singular values suggest that the quality of the
eigenvalues is good, with a few outliers.
%
These outliers are from the interval $[4.5,5]$.
%
In the plots, we observe a different type of
localization property, with many of the eigenfunctions
approximately supported in a small, connected subset
of the domain. This is consistent with other studies
\note{cite}

@@

>>>>

In this section, we demonstrate the analytical
claims above with numerical examples and
highlight the performance of the BIE approach
to Stokes eigenvalues.

@@

First, we describe the numerical tools needed
to compute Stokes eigenvalues in a BIE
framework.

@@

We can obtain an a posteriori estimate of the
error in a computed root as follows.
%
Let $f$ denote an analytic function,
$P$ be the polynomial interpolant
of that function over some interval,
$\delta f = f-P$ be the difference,
and $k_\cheb$ denote a computed root
of $P$.
%
The algorithm used by \texttt{chebfun}
to approximate the roots of $P$ is
backward stable~\cite{noferini2017chebyshev}.
Therefore the error in the roots will be
small relative to the error of the fit and
we set $P(k_\cheb) = 0$ below.
%
Suppose that
$f(k_\cheb + \delta k) = 0$ for some
small $\delta k$. Then

\begin{align*}
  0 &= f(k_\cheb + \delta k) \\
  0 &= P(k_\cheb + \delta k) + \delta f(k_\cheb + \delta k) \\
  \delta k &= -\frac{\delta f(k_\cheb + \delta k)}{P'(k_\cheb)} + O(\delta k^2) .
\end{align*}
In practice, we can obtain an approximate upper
bound for $|\delta f(k_\cheb+\delta k)|$
as $\epscheb \|P\|_\infty$ so that $\epscheb \|P\|_\infty/
|P'(k_\cheb)|$ provides an approximate upper bound
for the error in the root.

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_barbell_001_sings_plot}
    \caption{Smallest singular value of the BIE operator.}
    \label{subfig:barbell_sings}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_barbell_001_coeffs_plot}
    \caption{Normalized Chebyshev coefficients of the
      determinant on 3 intervals in $k$.}
    \label{subfig:barbell_coeffs}
  \end{subfigure}
  \caption{Diagnostics for the first 119 barbell eigenvalues.}
  \label{fig:barbell_diagnostics}
\end{figure}

We consider the barbell-shaped domain in \cref{fig:barbell}.
This domain is the union of a square of side-length 6,
a square of side-length 3, and a ``bridge'' connecting
them of height 1 and width 5/2.
%
For the sake of simplicity, we round the corners of the domain
to obtain a smooth object.
%
Applying the approach described in~\cite{epstein2016smoothed},
the corners of the domain are rounded by convolving with
the Gaussian kernel
\begin{equation}
  \nonumber
\phi(x) = \frac{1}{\sqrt{2\pi h}} e^{-x^2/(2 h^2)} \, ,
\end{equation}
with $h\approx 0.06$. This leaves the domain unperturbed
to high precision outside of a radius of $0.1$ around
each corner.
%
The eigenfunctions of such a domain display the well-known
localization property~\cite{trefethen2006computed}:
many of the eigenfunctions are approximately supported
within one of the squares.
%
We compute these eigenfunctions corresponding to
eigenvalues $k^2$ with $k$ in the range
$0.5 \leq k \leq 6.5$.

%
The panels are divided adaptively so that the smallest
panels in the rounded corners are smaller than $10^{-2}$,
which keeps the panels relatively flat.
%
This results in $N_p = 412$ after enforcing the
level-restriction property described in
\cref{rmk:levelrestrict}
and enforcing that no panel is larger than
one wavelength for the largest $k$
(here $\lambda=2\pi/6.5$).

As this is a simply-connected domain,
the eigenvalues are estimated by finding the values
$k$ for which $\cI-2\cDk-2\cW$ is non-invertible.
%
Let $f^N(k) = \det (\cI^N-2\cDk^N-2\cW^N)$.
To find the roots of $f^N(k)$, we fit a \texttt{chebfun}
representation of $f^N(k)$ on each of the intervals
$[j/2,(j+1)/2]$ for $j = 1,\ldots,12$.
%
We plot the absolute value of the Chebyshev coefficients
(normalized by the absolute value of the first coefficient)
of $f^N(k)$ on the intervals $[0.5,1.0]$, $[3.0,3.5]$,
and $[6.0,6.5]$ in \cref{subfig:barbell_coeffs}.
%
As expected, the coefficients decay exponentially
to zero, with more terms required at higher
frequencies.
%

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig/many_inclusions_gallery}
  \caption{Vorticity plots of the eigenfunctions corresponding
  to the first 76 eigenvalues of a domain with several inclusions.}
  \label{fig:many_inclusions_gallery}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_sings_plot}
    \caption{Smallest singular value of the BIE operator
      corresponding to the first 76 eigenvalues of a
      domain with several inclusions.}
    \label{subfig:many_inclusions_sings}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_coeffs_plot}
    \caption{Normalized Chebyshev coefficients of $f^N(k)$ on
      3 different intervals in $k$.}
    \label{subfig:many_inclusions_coeffs}
  \end{subfigure}
  \caption{Diagnostics for the eigenvalues of a domain
    with several inclusions.}
  \label{fig:many_inclusions_diagnostics}
\end{figure}

@@

%
For this smooth shape, ensuring that no panel is
larger than one wavelength for the largest $k$
(here $\lambda=2\pi/9$) is sufficient to resolve
the object to high precision.
%
After enforcing the
level-restriction property described in
\cref{rmk:levelrestrict}, we end up with
$N_p = 224$.

@@

We compute the roots of these Chebyshev interpolants
and apply the post-processing described above.
%
There were 103 total roots: 21 were removed because
the imaginary part was too large and 6 pairs were found
with values within $\sqrt{\epscheb}$ of each other.
%
For these 6 pairs, none represented two distinct
eigenvalues or a double root.
%
This leaves 76 roots in the range $0\leq k \leq 9$.
%
We plot the smallest singular value of
$\cI^N-\cDk^N-2i\cSk^N-2\cW^N$ for each of these roots in
\cref{subfig:many_inclusions_sings}
and plot the vorticity of the eigenfunctions
in \cref{fig:many_inclusions_gallery}.
%
In the vorticity plots, we observe a different type of
localization property than that seen in the barbell,
with many of the eigenfunctions
approximately supported in a small, connected subset
of the domain. This is consistent with other studies
\note{cite}

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_sings_plot_west}
    \caption{Smallest singular value of the BIE operator
      for the computed roots (original intervals)
      along with an estimate of the error in the root (dotted).}
    \label{subfig:many_inclusions_sings_west}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fig/ex_many_holes_004_sings_plot_ref}
    \caption{Smallest singular value of the BIE operator
      for the computed roots on the original intervals (\textbullet)
      and the values obtained on the refined interval ($\times$).}
    \label{subfig:many_inclusions_sings_ref}
  \end{subfigure}
  \caption{Further diagnostics for the eigenvalues of a domain
    with several inclusions.}
  \label{fig:many_inclusions_diagnostics_2}
\end{figure}

The singular values suggest that the quality of the
eigenvalues is good, with a few outliers.
%
To explain these outliers, we consider two quantities
which affect the singular value at a computed root.
%
As described above, we can approximate the
error in the computed root at $k_\cheb$ by
$\epscheb \|P\|_\infty/|P'(k_\cheb)|$,
where $P$ is the interpolating polynomial.
%
The singular value estimate itself is affected by
the error incurred in applying the inverse of the
compressed BIE matrix, which can be hard to quantify
\cite{ho2012fast}.
%
We approximate this error by $O(\sqrt{N})\epsflam$
and assume this is the order of the error in
the singular value estimate.
We plot the maximum of these two estimates
along with the computed singular values in
\cref{subfig:many_inclusions_sings_west}.
There is a reasonably good correlation between
the maximum of the error estimates and the
observed smallest singular value for the BIE,
especially for larger errors.

The worst outliers are from the left half
of the interval $[4.5,5]$.
%
Because the determinant is much larger on the
right half than the left half of $[4.5,5]$,
we can improve the estimate for the error
in the roots by subdividing the interval. 
We plot the smallest singular value of the
BIE for the roots obtained by fitting a polynomial
on $[4.5,4.75]$ in \cref{subfig:many_inclusions_sings_ref};
the roots on the refined interval
are of significantly higher quality.

\begin{remark}
  The above experience suggests that
  the ratio $\|P\|_\infty/|P'(k_\cheb)|$
  is a useful diagnostic for performing
  automated eigenvalue estimation.
  Note that at a multiple root, this ratio
  will be more difficult to bound.
\end{remark}
%
